Assessing the potential of Jacobian-free Newton-Krylov methods for cell-centred finite volume solid mechanics

### 2024-12-31 article.tex
Ready for co-author review

\author*[1,2,3]{\fnm{Philip} \sur{Cardiff}}\email{philip.cardiff@ucd.ie}
\author[4]{\fnm{Ivan} \sur{Batisti\'{c}}}
\author[4]{\fnm{\v{Z}eljko} \sur{Tukovi\'{c}}}

\affil*[1]{\orgdiv{School of Mechanical and Materials Engineering}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[2]{\orgdiv{UCD Centre for Mechanics}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[3]{\orgdiv{SFI I-Form Centre}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[4]{\orgdiv{Faculty of Mechanical Engineering and Naval Architecture}, \orgname{University of Zagreb}, \orgaddress{\country{Croatia}}}

@@

They found that the lower-order compact stencil preconditioner gave the best performance when combined with the GMRES linear solver and ILU(N) preconditioner.

@@

They also found a multigrid approach to be more efficient than ILU(N) for preconditioning, where the implementation was based on the PETSc \citep{PETSc} package.

@@

GMRES was the adopted linear solver with the ILU(N) preconditioner and fill-in values (N) ranging from 2 to 4. 

@@

%ILU(k) with k = 1-3. Approximate Jacobian.

@@

The governing equations are complemented by boundary conditions, with three types considered here: prescribed displacement, prescribed traction, and symmetry.
The definition of the engineering stress ($\bb{\sigma}_s$) and true stress ($\bb{\sigma}$) in Equations \ref{eqn:momentum_lingeom}, \ref{eqn:momentum_TL} and \ref{eqn:momentum_UL} is given by a chosen mechanical law.
Five mechanical laws are considered in this work, as briefly outlined in Appendix \ref{app:mechLaws}: linear elasticity (Hooke's law), St.\,Venant-Kirchhoff hyperelasticity, neo-Hookean hyperelasticity, Guccione hyperelasticity, and neo-Hookean $J_2$ hyperelastoplasticity.

@@

In the current work, we are interested in vector problems, where the governing momentum equation is formulated in terms of the unknown displacement solution vector.
In this case, Equation \ref{eq:J} refers to the individual scalar components of the residual, solution, and Jacobian.
That is, for 3-D analyses, the residual takes the form
\begin{eqnarray}
	\bb{R}(\bb{u}) = \left\{ R_1^x, R_1^y, R_1^z, R_2^x, R_2^y, R_2^z, ..., R_n^z \right\}
\end{eqnarray}
and the solution takes the form
\begin{eqnarray}
	\bb{u} = \left\{ u_1^x, u_1^y, u_1^z, u_2^x, u_2^y, u_2^z, ..., u_n^z \right\}
\end{eqnarray}
In practice, it is often more practical and efficient to form and store the residual, solution and Jacobian in a \emph{blocked} manner, where the residual and solution can be considered as vectors of vectors.
Similarly, the Jacobian can be formed in terms of sub-matrix block coefficients.

@@

Within the square braces on thr right-hand side of Equation \ref{eq:RhieChow}, the first terms represent a compact stencil (two-node) approximation of the face normal gradient, while the second terms represent a larger stencil approximation.
These two terms cancel out in the limit of mesh refinement (or if the solution varies linearly); otherwise, they produce a stabilisation effect that tends to smooth the solution fields.
As the term reduces at a third-order \hl{or second?} rate, it does not affect the overall scheme's second-order accuracy; the 1-D local truncation error is derived for this Rhie-Chow stabilisation term in Appendix \ref{app:RhieChow}.

@@

The linearised system (Equation \ref{eq:Seg}) is formed for each cell in the domain, resulting in a system of algebraic equations:
\begin{eqnarray} \label{eq:SegSys}
    \left[ \bb{\tilde{J}} \right]  \; \left[ \delta \bb{u} \right] = - \left[\bb{R}(\bb{u}_n)\right]
\end{eqnarray}
where $\left[ \bb{\tilde{J}} \right]$ is a symmetric, $M \times M$ stiffness matrix, where $M = 3|\mathcal{P}|$ in 3-D and $M = 3|\mathcal{P}|$ in 2-D.
In addition, if $\Delta t < \infty$ or $\mathcal{F}^{\text{disp}}_P \neq \emptyset$, matrix $\left[ \bb{\tilde{J}} \right]$ is strongly diagonally dominant.
The block ($3\times3$ for 3-D, $2\times2$ for 2-D) diagonal coefficient for cell $P$ (row $P$, column $P$) can be expressed as
\begin{eqnarray}
	 \left[ \tilde{\bb{J}}\right]_{PP} &=&
		\left[
		- \sum_{f_i \in \mathcal{F}^{\text{int}}_P}  \bar{K}
		\frac{ \left|\bb{\Delta}_{f_i} \right| }{\left|\bb{d}_{f_i}\right|}    \left|\bb{\Gamma}_{f_i}\right| \textbf{I} 
	    \quad-  \sum_{d_i \in \mathcal{F}^{\text{disp}}_P}  \bar{K}
%		\left|\bb{\Delta}_{d_i} \right| \left( \ \frac{ \bar{\bb{u}}_{d_i} - \bb{u}_P}{\left|\bb{d}_{d_i}\right|}  \right) 
		 \frac{ \left|\bb{\Delta}_{d_i} \right| }{\left|\bb{d}_{d_i}\right|} 
		\left|\bb{\Gamma}_{d_i}\right| \textbf{I} \right. \notag \\
	 &&\quad - \left. \sum_{s_i \in \mathcal{F}^{\text{symm}}_P}  \bar{K}
		 \frac{ \left|\bb{\Delta}_{s_i} \right|}{\left|\bb{d}_{s_i}\right|}
		\left|\bb{\Gamma}_{s_i}\right|  \bb{n}_{s_i} \otimes \bb{n}_{s_i} 
	\quad - \quad \frac{9}{4}  \frac{\rho_P \Omega_P}{\Delta t^2} \textbf{I} \right]
\end{eqnarray}
while the off-diagonal coefficients (row $P$, column $Q$) can be expressed as
\begin{eqnarray}
	\left[\tilde{\bb{J}}\right] _{PQ} &=&
		\bar{K} \frac{ \left|\bb{\Delta}_{f_{PQ}} \right| }{\left|\bb{d}_{f_{PQ}}\right|}    \left|\bb{\Gamma}_{f_{PQ}}\right| \textbf{I} 
\end{eqnarray}
where subscript $PQ$ indicates a quantity associated with the internal face $f$ shared between cells $P$ and $Q$.
%where $N_f$ represents the set of faces $f$ in cell $P$, and neighbouring cell centre $N_f$ shares face $f$ with the cell $P$.
%The over-relaxed orthogonal vector $\bb{\Delta}_f = \frac{\bb{d}_f}{\bb{d}_f \cdot \bb{n}_f}$ 
%The non-orthogonal correction vector $\bb{k}_f=\bb{n}_f-\bb{\Delta}_f$.
% where $\bb{n}_f$ is the outward-facing unit normal to the face $f$.
%Vector $\bb{d}_f$ connects the centre of cell $P$ with the centre of cell $N_f$ in the updated configuration.
%The first term on the right-hand side is treated implicitly, while the second term - representing non-orthogonal corrections at the face - is treated in a deferred correction manner.

@@

By design, matrix $\left[ \bb{\tilde{J}} \right]$  contains no inter-component coupling; consequently, three equivalent smaller linear systems can be formed and solved for the Cartesian components of the displacement correction (or two in 2-D), e.g.
\begin{eqnarray} \label{eq:SegSysX}
%     \bb{\tilde{J}}_x(\bb{u}_n)  \;  \Delta \bb{u}_x = - \mathcal{R}_x(\bb{u}_n) \label{eq:segX} \\
%     \bb{\tilde{J}}_y(\bb{u}_n)  \;  \Delta \bb{u}_y = - \mathcal{R}_y(\bb{u}_n) \label{eq:segY} \\
%     \bb{\tilde{J}}_z(\bb{u}_n)  \;  \Delta \bb{u}_z = - \mathcal{R}_z(\bb{u}_n) \label{eq:segZ}
     \left[ \bb{\tilde{J}}_x \right]  \;  \left[ \Delta \bb{u}_x \right] = - \left[ \mathcal{R}_x(\bb{u}_n) \right] \label{eq:segX} \\
     \left[ \bb{\tilde{J}}_y \right]  \;  \left[ \Delta \bb{u}_y \right] = - \left[ \mathcal{R}_y(\bb{u}_n) \right] \label{eq:segY} \\
     \left[ \bb{\tilde{J}}_z \right]  \;  \left[ \Delta \bb{u}_z \right] = - \left[ \mathcal{R}_z(\bb{u}_n) \right] \label{eq:segZ}
\end{eqnarray}
where $ \bullet_x$ represents the components in the $x$ direction, $ \bullet_y$ represents the components in the $y$ direction, and $ \bullet_z$ represents the components in the $z$ direction.
Matrices $\left[ \bb{\tilde{J}}_x \right]$, $ \left[ \bb{\tilde{J}}_y\right]$ and $\left[ \bb{\tilde{J}}_z\right]$ have the size $|\mathcal{P}| \times |\mathcal{P}|$.
An additional benefit from a memory perspective, is that matrices $\left[ \bb{\tilde{J}}_x \right]$, $ \left[ \bb{\tilde{J}}_y\right]$ and $\left[ \bb{\tilde{J}}_z\right]$ are identical, except for the effects from including boundary conditions ($\mathcal{F}^{\text{symm}}_P$ terms).
From an implementation perspective, this allows a single scalar matrix to be formed and stored, where the boundary condition contributions are inserted before solving a particular component.

The \emph{inner} linear sparse systems (Equations \ref{eq:segX}, \ref{eq:segY} and \ref{eq:segZ}) can be solved using any typical direct or iterative linear solver approach; however, an incomplete Cholesky pre-conditioned conjugate gradient method \cite{Jacobs1986} is often preferred as the diagonally dominant characteristic leads to good convergence characteristics.
Algebraic multigrid can be used to accelerate convergence.
%In non-linear problems, this system of equations is solved multiple times with updated coefficients in a fixed-point iteration scheme. 
%As noted in previous articles on segregated methods, the inner system need not be solved to a tight tolerance as coefficients and source terms are approximated from the previous increment; a reduction in the residuals of one order of magnitude is typically sufficient. The outer iterations are performed until the predefined tolerance, typically $1 \times 10^{-6}$, has been achieved \cite{cardiff_lagrangian_2017}. 
%In the current updated Lagrangian approach, the mesh is moved to the deformed configuration at the end of each time step rather than after each outer iteration.
%Since the displacements are calculated at the cell centres, a linear least-squared method is employed here \cite{cardiff_lagrangian_2017} to interpolate the displacement increments to the mesh vertices, allowing the mesh to be moved.
%In this method, a linear least squares plane is fit through a vertex and its immediately adjacent cell centres. For boundary vertices, boundary face-centre values are also included in the fitting.
%
%The procedures have been implemented and publicly shared within the solids4foam toolbox \citep{Cardiff2018, Tukovic2018} of the open-source OpenFOAM software.

@@

\hl{Comment: we can solve 31 or 34-36; we are using 31 in PETSc, while the native OF code uses 34-36}
\hl{We should comment on this}
%The current procedure is implemented and publicly shared in the solids4foam toolbox of OpenFOAM.
%\hl{Add a section about code sharing: appendix?}
%\hl{Comment: we have two implementations of segregated: native OpenFOAM (solves Eqs 16-18) and PETSc SNES (solves Eq 15)}
%\hl{Do we need to comment on this? Maybe we should use only PETSc SNES for a fair comparison}
%\hl{Or we could use both on the verification case and then stick with just one afterwards}

@@

As noted in the introduction, the Jacobian-free Newton-Krylov avoids the need to explicitly construct the Jacobian matrix by approximating its action on a solution vector using the finite difference method, repeated here:
\begin{eqnarray} \label{eq:JF}
	\bb{J} \bb{v} \approx \frac{\bb{F}(\bb{x} + \epsilon \bb{v}) - \bb{F}(\bb{x})}{\epsilon}
\end{eqnarray}

@@

%The literature indicates that the choice of preconditioner for the inner linearised system has a major impact on the efficiency and robustness of the overall solution procedure.
The purpose of preconditioning the Jacobian-free Newton-Krylov method is to reduce the number of inner linear solver iterations.
In the current work, the GMRES linear solver is used for the inner system.
%Left or right preconditioning, may be employed in a Jacobian-free context, and there are pros and cons to both.
Using right preconditioning, the finite difference approximation of Equation \ref{eq:JF} becomes
\begin{eqnarray}
%	(\bb{J} \bb{P}^{-1}) (\bb{P} \delta \bb{u}) = -\bb{R}(\bb{u})
	\bb{J} \bb{P}^{-1} \bb{v}
	\approx
	\frac{\bb{F}(\bb{x} + \epsilon \bb{P}^{-1} \bb{v}) - \bb{F}(\bb{x})}{\epsilon}
\end{eqnarray}
where $\bb{P}$ is the preconditioning matrix or process.
In practice, only the action of $\bb{P}^{-1}$ on a vector is required, and the $\bb{P}^{-1}$ may not be explicitly formed.
Concretely, the preconditioner needs to approximately solve the linear system $\bb{y} = \bb{P}^{-1} \bb{v}$.
%Thus, while we may refer to the matrix P, operationally the algorithm only requires the action of $P^{-1}$ on a vector.

%\hl{3}
%- we use approx J to form precon
%- many precon used in lit, we will consider ILU(N) and MG, where MG is expected to be better, but also LU since direct solvers are popular in FE solid mechanics
In the current work, we propose to use the compact-stencil approximate Jacobian from the segregated algorithm $\tilde{\bb{J}}$ as the preconditioning matrix $\bb{P}$ for the preconditioned Jacobian-free Newton-Krylov method.
%\hl{Comment on literature: used before but not yet for solids}.
This preconditioning approach can be considered as a ``physics-based" preconditioner in the classifications of \citet{Knoll2004}.
The approach is conceptually similar to an approximation of the Jacobian of a higher-order large-stencil scheme by a compact-stencil lower-order scheme.
A benefit of the proposed approach is that existing segregated frameworks can re-use their existing discretisation and storage implementations.
Concretely, the Jacobian-free Newton-Krylov method requires only a procedure for forming this preconditioning matrix and a procedure for explicitly evaluating the residual.
Both routines are easily implemented -- and are likely already available -- in an existing segregated framework.
The only additional required procedure is an interface to an existing Jacobian-free Newton-Krylov implementation.
%"The motivation behind this approach is that there exist numerous, legacy algorithms to solve nonlinear systems, both IVPs and BVPs. These algorithms typically were developed with some insight into the time scales or physical behavior of the problem. As a benefit of this insight, a reduced implicit system, or a sequence of segregated explicit or implicit systems may be solved in place of the fully coupled system. "
In the current work, the PETSc package \cite{PETSc} is used as the nonlinear solver, driven by a finite volume solver in the solids4foam toolbox \citep{Cardiff2018, Tukovic2018a} for OpenFOAM toolbox \citep{Weller1998} (version OpenFOAM-v2312).

Several preconditioners are available in the literature, incomplete Cholesky or ILU(k) being popular; however, multigrid methods offer the greatest potential for large-scale problems.
As noted by \citet{Knoll2004}, algorithmic simplifications within a multigrid procedure, which may result in loss of convergence for multigrid as a solver, have a much weaker effect when multigrid is used as a preconditioner.
In this work, three preconditioners are considered:
\begin{enumerate}
	\item ILU(k): incomplete lower-upper decomposition with fill-in k. %The segregated solver uses ILU(0), that is, ILU with zero fill-in.
	\item Multigrid: the HYPRE Boomerang \citep{hypre} multigrid implementation.
	\item LU: the MUMPS \citep{MUMPS:1, MUMPS:2} lower-upper decomposition direct solver.
\end{enumerate}

>>>>

\author*[1,2,3]{\fnm{Philip} \sur{Cardiff}}\email{philip.cardiff@ucd.ie}
\author[4]{\fnm{Ivan} \sur{Batisti\'{c}}}
\author[5]{\fnm{Hiroaki} \sur{Nishikawa}}
\author[4]{\fnm{\v{Z}eljko} \sur{Tukovi\'{c}}}

\affil*[1]{\orgdiv{School of Mechanical and Materials Engineering}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[2]{\orgdiv{UCD Centre for Mechanics}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[3]{\orgdiv{SFI I-Form Centre}, \orgname{University College Dublin}, \orgaddress{\country{Ireland}}}
\affil[4]{\orgdiv{Faculty of Mechanical Engineering and Naval Architecture}, \orgname{University of Zagreb}, \orgaddress{\country{Croatia}}}
\affil[5]{\orgname{National Institute of Aerospace}, \orgaddress{\country{Hampton, VA 23666, USA}}}

@@

They found that the lower-order compact stencil preconditioner gave the best performance when combined with the GMRES linear solver and ILU($k$) preconditioner.

@@

They also found a multigrid approach to be more efficient than ILU($k$) for preconditioning, where the implementation was based on the PETSc \citep{PETSc} package.

@@

GMRES was the adopted linear solver with the ILU($k$) preconditioner and fill-in values ($k$) ranging from 2 to 4. 

@@

%ILU($k$) with k = 1-3. Approximate Jacobian.

@@

The governing equations are complemented by boundary conditions, with three types considered here: prescribed displacement, prescribed traction, and symmetry.
The definition of the engineering stress ($\bb{\sigma}_s$) and true stress ($\bb{\sigma}$) in Equations \ref{eqn:momentum_lingeom}, \ref{eqn:momentum_TL} and \ref{eqn:momentum_UL} is given by a chosen mechanical law.
Five mechanical laws are considered in this work, as briefly outlined in Appendix \ref{app:mechLaws}: linear elasticity (Hooke's law), three forms of hyperelasticity (St.\,Venant-Kirchhoff, neo-Hookean, and Guccione), and neo-Hookean $J_2$ hyperelastoplasticity.

@@

The current work focusses on vector problems, where the governing momentum equation is formulated in terms of the unknown displacement solution vector.
In this case, Equation \ref{eq:J} refers to the individual scalar components of the residual, solution, and Jacobian.
That is, for 3-D analyses, the residual takes the form
\begin{eqnarray}
	\bb{R}(\bb{u}) = \left\{ R_1^x, R_1^y, R_1^z, R_2^x, R_2^y, R_2^z, ..., R_n^z \right\}
\end{eqnarray}
and the solution takes the form
\begin{eqnarray}
	\bb{u} = \left\{ u_1^x, u_1^y, u_1^z, u_2^x, u_2^y, u_2^z, ..., u_n^z \right\}
\end{eqnarray}
In practice, it is often more practical and efficient to form and store the residual, solution and Jacobian in a \emph{blocked} manner, where the residual and solution can be considered as vectors of vectors.
Similarly, the Jacobian can be formed in terms of sub-matrix block coefficients.

@@

Within the square braces on thr right-hand side of Equation \ref{eq:RhieChow}, the first terms represent a compact stencil (two-node) approximation of the face normal gradient, while the second terms represent a larger stencil approximation.
These two terms cancel out in the limit of mesh refinement (or if the solution varies linearly); otherwise, they produce a stabilisation effect that tends to smooth the solution fields.
The magnitude of the stabilisation is shown in Appendix \ref{app:RhieChow} to reduce at a second-order rate (not a third-order rate as stated previously \citep{Demirdzic1995}), and hence does not affect the overall scheme's second-order accuracy.

@@

The linearised system (Equation \ref{eq:Seg}) is formed for each cell in the domain, resulting in a system of algebraic equations:
\begin{eqnarray} \label{eq:SegSys}
    \left[ \bb{\tilde{J}} \right]  \; \left[ \delta \bb{u} \right] = - \left[\bb{R}(\bb{u}_n)\right]
\end{eqnarray}
where $\left[ \bb{\tilde{J}} \right]$ is a symmetric, $M \times M$ stiffness matrix, where $M = 3|\mathcal{P}|$ in 3-D and $M = 2|\mathcal{P}|$ in 2-D.
If $\Delta t < \infty$ or $\mathcal{F}^{\text{disp}}_P \neq \emptyset$, matrix $\left[ \bb{\tilde{J}} \right]$ is strongly diagonally dominant; otherwise, it is weakly diagonally dominant.
The block ($3\times3$ for 3-D, $2\times2$ for 2-D) diagonal coefficient for cell $P$ (row $P$, column $P$) can be expressed as
\begin{eqnarray}
	 \left[ \tilde{\bb{J}}\right]_{PP} &=&
		- \sum_{f_i \in \mathcal{F}^{\text{int}}_P}  \bar{K}
		\frac{ \left|\bb{\Delta}_{f_i} \right| }{\left|\bb{d}_{f_i}\right|}    \left|\bb{\Gamma}_{f_i}\right| \textbf{I} 
	    \quad-  \sum_{d_i \in \mathcal{F}^{\text{disp}}_P}  \bar{K}
%		\left|\bb{\Delta}_{d_i} \right| \left( \ \frac{ \bar{\bb{u}}_{d_i} - \bb{u}_P}{\left|\bb{d}_{d_i}\right|}  \right) 
		 \frac{ \left|\bb{\Delta}_{d_i} \right| }{\left|\bb{d}_{d_i}\right|} 
		\left|\bb{\Gamma}_{d_i}\right| \textbf{I} \notag \\
	 &&\quad - \sum_{s_i \in \mathcal{F}^{\text{symm}}_P}  \bar{K}
		 \frac{ \left|\bb{\Delta}_{s_i} \right|}{\left|\bb{d}_{s_i}\right|}
		\left|\bb{\Gamma}_{s_i}\right|  \bb{n}_{s_i} \otimes \bb{n}_{s_i} 
	\quad - \quad \frac{9}{4}  \frac{\rho_P \Omega_P}{\Delta t^2} \textbf{I}
\end{eqnarray}
while the off-diagonal coefficients (row $P$, column $Q$) can be expressed as
\begin{eqnarray}
	\left[\tilde{\bb{J}}\right] _{PQ} &=&
		\bar{K} \frac{ \left|\bb{\Delta}_{f_{PQ}} \right| }{\left|\bb{d}_{f_{PQ}}\right|}    \left|\bb{\Gamma}_{f_{PQ}}\right| \textbf{I} 
\end{eqnarray}
where subscript $PQ$ indicates a quantity associated with the internal face $f$ shared between cells $P$ and $Q$.
%where $N_f$ represents the set of faces $f$ in cell $P$, and neighbouring cell centre $N_f$ shares face $f$ with the cell $P$.
%The over-relaxed orthogonal vector $\bb{\Delta}_f = \frac{\bb{d}_f}{\bb{d}_f \cdot \bb{n}_f}$ 
%The non-orthogonal correction vector $\bb{k}_f=\bb{n}_f-\bb{\Delta}_f$.
% where $\bb{n}_f$ is the outward-facing unit normal to the face $f$.
%Vector $\bb{d}_f$ connects the centre of cell $P$ with the centre of cell $N_f$ in the updated configuration.
%The first term on the right-hand side is treated implicitly, while the second term - representing non-orthogonal corrections at the face - is treated in a deferred correction manner.

@@

By design, matrix $\left[ \bb{\tilde{J}} \right]$  contains no inter-component coupling, i.e., each block coefficient is diagonal; consequently, three equivalent smaller linear systems can be formed in 3-D (or two linear systems in 2-D) and solved for the Cartesian components of the displacement correction, e.g.
\begin{eqnarray} \label{eq:SegSysX}
%     \bb{\tilde{J}}_x(\bb{u}_n)  \;  \Delta \bb{u}_x = - \mathcal{R}_x(\bb{u}_n) \label{eq:segX} \\
%     \bb{\tilde{J}}_y(\bb{u}_n)  \;  \Delta \bb{u}_y = - \mathcal{R}_y(\bb{u}_n) \label{eq:segY} \\
%     \bb{\tilde{J}}_z(\bb{u}_n)  \;  \Delta \bb{u}_z = - \mathcal{R}_z(\bb{u}_n) \label{eq:segZ}
     \left[ \bb{\tilde{J}}_x \right]  \;  \left[ \Delta \bb{u}_x \right] = - \left[ \mathcal{R}_x(\bb{u}_n) \right] \label{eq:segX} \\
     \left[ \bb{\tilde{J}}_y \right]  \;  \left[ \Delta \bb{u}_y \right] = - \left[ \mathcal{R}_y(\bb{u}_n) \right] \label{eq:segY} \\
     \left[ \bb{\tilde{J}}_z \right]  \;  \left[ \Delta \bb{u}_z \right] = - \left[ \mathcal{R}_z(\bb{u}_n) \right] \label{eq:segZ}
\end{eqnarray}
where $ \bullet_x$ represents the components in the $x$ direction, $ \bullet_y$ represents the components in the $y$ direction, and $ \bullet_z$ represents the components in the $z$ direction.
Matrices $\left[ \bb{\tilde{J}}_x \right]$, $ \left[ \bb{\tilde{J}}_y\right]$ and $\left[ \bb{\tilde{J}}_z\right]$ have the size $|\mathcal{P}| \times |\mathcal{P}|$.
An additional benefit from a memory and assembly perspective, is that matrices $\left[ \bb{\tilde{J}}_x \right]$, $ \left[ \bb{\tilde{J}}_y\right]$ and $\left[ \bb{\tilde{J}}_z\right]$ are identical, except for the effects from including boundary conditions ($\mathcal{F}^{\text{symm}}_P$ terms).
From an implementation perspective, this allows a single scalar matrix to be formed and stored, where the boundary condition contributions are inserted before solving a particular component.

The \emph{inner} linear sparse systems (Equations \ref{eq:segX}, \ref{eq:segY} and \ref{eq:segZ}) can be solved using any typical direct or iterative linear solver approach; however, an incomplete Cholesky pre-conditioned conjugate gradient method \cite{Jacobs1986} is often preferred as the diagonally dominant characteristic leads to good convergence characteristics.
Algebraic multigrid can also be used to accelerate convergence.
%In non-linear problems, this system of equations is solved multiple times with updated coefficients in a fixed-point iteration scheme. 
%As noted in previous articles on segregated methods, the inner system need not be solved to a tight tolerance as coefficients and source terms are approximated from the previous increment; a reduction in the residuals of one order of magnitude is typically sufficient. The outer iterations are performed until the predefined tolerance, typically $1 \times 10^{-6}$, has been achieved \cite{cardiff_lagrangian_2017}. 
%In the current updated Lagrangian approach, the mesh is moved to the deformed configuration at the end of each time step rather than after each outer iteration.
%Since the displacements are calculated at the cell centres, a linear least-squared method is employed here \cite{cardiff_lagrangian_2017} to interpolate the displacement increments to the mesh vertices, allowing the mesh to be moved.
%In this method, a linear least squares plane is fit through a vertex and its immediately adjacent cell centres. For boundary vertices, boundary face-centre values are also included in the fitting.
%
%The procedures have been implemented and publicly shared within the solids4foam toolbox \citep{Cardiff2018, Tukovic2018} of the open-source OpenFOAM software.

@@

%\hl{Comment: we can solve 31 or 34-36; we are using 31 in PETSc, while the native OF code uses 34-36}
%\hl{We should comment on this}
%The current procedure is implemented and publicly shared in the solids4foam toolbox of OpenFOAM.
%\hl{Add a section about code sharing: appendix?}
%\hl{Comment: we have two implementations of segregated: native OpenFOAM (solves Eqs 16-18) and PETSc SNES (solves Eq 15)}
%\hl{Do we need to comment on this? Maybe we should use only PETSc SNES for a fair comparison}
%\hl{Or we could use both on the verification case and then stick with just one afterwards}

@@

As noted in the introduction, the Jacobian-free Newton-Krylov method avoids the need to explicitly construct the Jacobian matrix by approximating its action on a solution vector using the finite difference method, repeated here:
\begin{eqnarray} \label{eq:JF}
	\bb{J} \bb{v} \approx \frac{\bb{F}(\bb{x} + \epsilon \bb{v}) - \bb{F}(\bb{x})}{\epsilon}
\end{eqnarray}

@@

%The literature indicates that the choice of preconditioner for the inner linearised system has a major impact on the efficiency and robustness of the overall solution procedure.
The purpose of preconditioning the Jacobian-free Newton-Krylov method is to reduce the number of inner linear solver iterations.
In the current work, the GMRES linear solver is used for the inner system.
%Left or right preconditioning, may be employed in a Jacobian-free context, and there are pros and cons to both.
Using right preconditioning, the finite difference approximation of Equation \ref{eq:JF} becomes
\begin{eqnarray}
%	(\bb{J} \bb{P}^{-1}) (\bb{P} \delta \bb{u}) = -\bb{R}(\bb{u})
	\bb{J} \bb{P}^{-1} \bb{v}
	\approx
	\frac{\bb{F}(\bb{x} + \epsilon \bb{P}^{-1} \bb{v}) - \bb{F}(\bb{x})}{\epsilon}
\end{eqnarray}
where $\bb{P}$ is the preconditioning matrix or process.
In practice, only the action of $\bb{P}^{-1}$ on a vector is required, and $\bb{P}^{-1}$ need not be explicitly formed.
Concretely, the preconditioner needs to approximately solve the linear system $\bb{y} = \bb{P}^{-1} \bb{v}$.
%Thus, while we may refer to the matrix P, operationally the algorithm only requires the action of $P^{-1}$ on a vector.

%\hl{3}
%- we use approx J to form precon
%- many precon used in lit, we will consider ILU($k$) and MG, where MG is expected to be better, but also LU since direct solvers are popular in FE solid mechanics
In the current work, we propose to use the compact-stencil approximate Jacobian from the segregated algorithm $\tilde{\bb{J}}$ as the preconditioning matrix $\bb{P}$ for the preconditioned Jacobian-free Newton-Krylov method.
%\hl{Comment on literature: used before but not yet for solids}.
This preconditioning approach can be considered as a ``physics-based" preconditioner in the classifications of \citet{Knoll2004}.
The approach is conceptually similar to an approximation of the Jacobian of a higher-order large-stencil scheme by a compact-stencil lower-order scheme.
A benefit of the proposed approach is that existing segregated frameworks can re-use their existing matrix assembly and storage implementations.
Concretely, the Jacobian-free Newton-Krylov method requires only a procedure for forming this preconditioning matrix and a procedure for explicitly evaluating the residual.
Both routines are easily implemented -- and are likely already available -- in an existing segregated framework.
The only additional required procedure is an interface to an existing Jacobian-free Newton-Krylov implementation.
%"The motivation behind this approach is that there exist numerous, legacy algorithms to solve nonlinear systems, both IVPs and BVPs. These algorithms typically were developed with some insight into the time scales or physical behavior of the problem. As a benefit of this insight, a reduced implicit system, or a sequence of segregated explicit or implicit systems may be solved in place of the fully coupled system. "
In the current work, the PETSc package \cite{PETSc} is used as the nonlinear solver, driven by a finite volume solver implemented in the solids4foam toolbox \citep{Cardiff2018, Tukovic2018a} for OpenFOAM toolbox \citep{Weller1998} (version OpenFOAM-v2312).

Several preconditioners are available in the literature, incomplete Cholesky or ILU($k$) being popular; however, multigrid methods offer the greatest potential for large-scale problems.
As noted by \citet{Knoll2004}, algorithmic simplifications within a multigrid procedure, which may result in loss of convergence for multigrid as a solver, have a much weaker effect when multigrid is used as a preconditioner.
In this work, three preconditioners are considered:
\begin{enumerate}
	\item ILU($k$): incomplete lower-upper decomposition with fill-in $k$. %The segregated solver uses ILU(0), that is, ILU with zero fill-in.
	\item Multigrid: the HYPRE Boomerang \citep{hypre} multigrid implementation.
	\item LU: the MUMPS \citep{MUMPS:1, MUMPS:2} lower-upper decomposition direct solver.
\end{enumerate}

%\hl{4 - globalisation}
A challenge with Newton-type methods, including Jacobian-free versions, is poor convergence when far from the true solution, and divergence is often a real possibility.
Globalisation refers to steering an initial solution towards the quadratic convergence range of the Newton method.
Several strategies are possible, and it is common to combine approaches \cite{Knoll2004}.
In the current work, a line search procedure is used to select the $s$ parameter in the solution update step (the second line in Equations \ref{eq:NewtonRaphson}).
Line search methods assume the Newton update direction is correct and aim to find a scalar $s > 0$ that decreases the residual $\bb{R}(\bb{u}_k + s \delta \bb{u}) < \bb{R}(\bb{u}_k)$.
%The scalar $s$ is typically $\leq 1$, but extrapolation ($>1$) is also possible for accelerating convergence, albeit at the expense of robustness.
In addition to a line search approach, a \emph{transient continuation} globalisation approach is used in the current work, where the displacement $\bb{u}_P$ for cell $P$ at time $t + \Delta t$ is predicted at the start of a new time (or loading) step, based on a truncated second-order Taylor series expansion:
\begin{eqnarray} \label{eq:predictor}
%	\bb{u}^{[t+\Delta t]} = \bb{u}^{[t]} + \Delta t \left(\frac{\partial \bb{u}}{\partial t}\right)^{[t]} + \frac{1}{2} \Delta t^2 \left( \frac{\partial^2 \bb{u}}{\partial t^2} \right)^{[t]}
	\bb{u}^{[t+\Delta t]}_P = \bb{u}^{[t]}_P + \Delta t \, \bb{v}^{[t]}_P + \frac{1}{2} \Delta t^2 \left( \frac{\partial \bb{v}}{\partial t} \right)^{[t]}_P
\end{eqnarray}
where $\bb{v}^{[t]}_P$ is the velocity of cell $P$ at time $t$, and $\left( \frac{\partial \bb{v}}{\partial t} \right)^{[t]}_P$ is the acceleration.
%$\left( \frac{\partial^2 \bb{u}}{\partial t^2} \right)_t$ 
In this way, for highly nonlinear problems, the user can decrease the time step size $\Delta t$ as a globalisation approach to improve the performance of the Newton method.
The predictor step in Equation \ref{eq:predictor} has been chosen to be consistent with the discretisation of the inertia term (Equation \ref{eq:inertia2}).
